---
title: "wiki_development_notes"
---

# Overview

> Historical working notes. The implemented core API and decisions now live in
> [dev/api-specification.md](../api-specification.md). See the migration vignette
> for current syntax; the exploratory code below is retained as history and is
> excluded from the package build. Use testthat::test_local() for current tests.

- Working my way through the Wiki development notes (https://github.com/brad-cannell/freqtables/wiki/Development-Notes). 

- At this point, don't worry about great documentation. Just get things working. You can document them and add them to R scripts later.

- Having said that, do try to write modularized code -- break the main function up into smaller functions that can individually be maintained and tested.

- Delete this before submitting to CRAN.

## Process

- As the functions get longer, it's actually easier to to keep them in a separate R script and then source them than it is to scroll up and down running them and testing them.

- Start by making the functions you write accomplish the main task they are supposed to accomplish. Then, come back and put in data checks.

- While developing you will want to test the impact of changes you make to the function. 
  - You can run mini tests right in this Quarto document. Meaning write a little code chunk to test a specific thing.
  - You can also run saved tests directly in `test-freq_tbl.R`.
  - You can run all the tests in `test-freq_tbl.R` by running `source("tests/testthat/test-freq_tbl.R")`.

# Load packages

Don't load packages. Use the double colon syntax like you would in an R script. 

# Load data 

For testing

```{r}
data(freq_study)
```

# freq_tbl

- (Done) Finish creating unit tests: tests/testthat/test-freq_tbl.R. You can use test-freq_table.R for inspiration.

- (Done) Make sure the basics of the function are working as expected.
  
- (Done) Create an informative error when the value passed to the .freq_var argument isn't a column of .data.

- (Done) Create an informative error when nothing is passed to freq_tbl()
  - Currently, it just says "Error in as_string(x) : argument "x" is missing, with no default"

- Create an informative error for when more than one column name is passed to freq_tbl().
  - Currently, it just says "object 'sex' not found".
  
- Move all the tests from freq_tbl_test.R to test-freq_tbl.R
  - Maybe do this on the plane.

- We may want to come back and add a .col_headers = "col_names" argument that users can adjust if they want generic column headers instead.

- We can add an overall = FALSE argument that users who want an overall percentage can adjust (two-way and n-way tables).
  - Maybe. Now, I think the default should be: n, prop, n_group, prop_group
  - Annotate this in the wiki.
  
- We may want to add a special class to the resut of freq_tbl(), similar to what we did for freq_table().

- (Don't do this) When you get this working, try breaking the function up into smaller functions.
  - I tried playing with this. I don't think there's enough code there for this to make sense.
  - Breaking code up into functions is primariy to reduce repitition. There isn't really any repitition in the freq_tbl code.
  
  

Create an informative error when nothing is passed to freq_tbl()

```{r}
test <- function(.data, .freq_var, .drop = FALSE, .percent = FALSE) {
  drop_chr <- deparse(substitute(.drop))
  
  if (!(drop_chr %in% c(TRUE, FALSE))) {
    stop(glue::glue(
      "freq_tbl() expects the value passed to the '.freq_var' argument (the",
      " second argument) to be a column of the data frame passed to the",
      " '.data.' argument.",
      " The value currently being passed to '.freq_var' is called '{freq_var_chr}'",
      " and is not recognized as a column in the data frame being passed to",
      " '.data.'"
    ))
  }
}

freq_study |> test(exposure, sex)
```


```{r}
# Run all tests
devtools::load_all()
source("tests/testthat/test-freq_tbl.R")
```





















# Dumping ground

Stuff I think I don't need anymore, but I'm not 100% sure.

## freq_tbl_data_check

Make this a separate function for easier maintenance and testing. 

```{r}
freq_tbl_data_check <- function(.data, .freq_var) {
  # ===========================================================================
  # Prevents R CMD check: "no visible binding for global variable ‘.’"
  # ===========================================================================
  # Fill this in later.
  
  # ===========================================================================
  # Check to see what type of object is being passed to .data
  #   - Should be a data frame.
  #   - For example, some people use "attach". That won't work with freq_tbl()
  # ===========================================================================
  .data_class <- class(.data)
  if (!("data.frame" %in% .data_class)) {
    stop(
      paste0(
        "freq_tbl expects the object passed to the .data argument (the",
        " first argument) to be a data frame. Currently, the object being",
        " passed to .data has the class: ", .data_class, ". Please use the",
        " form mtcars |> freq_table(am) or freq_table(mtcars, am)."
      )
    )
  }
  
  # ===========================================================================
  # Check to make sure there was a value passed to the .freq_var argument.
  # And that the value is a column that exists in .data.
  # ===========================================================================
  if (is.null(.freq_var)) {
    stop(
      "It's empty."
    )
  }
  # Defuse the user expression in .freq_var
  .freq_var <- rlang::enquo(.freq_var)
  # out <- dplyr::pull(.data, rlang::enquo(.freq_var))
  return(.freq_var)
  
  # Pass .data along if it clears data checks
  .data
}
```

## Test freq_tbl_data_check

```{r}
# Not a data frame
# freq_study$sex |> freq_tbl_data_check() # Return informative error

# Nothing passed to .freq_var
# freq_study |> freq_tbl_data_check() # Return informative error
```


Actually, I'm not sure I want to do it this way. Maybe I should put the data checks inside the main function. 
- I want to check all of the arguments.
- I don't want to have to enquo arguments multiple times.
- Still break the stats functions into multiple functions when that makes sense.
