---
output:
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
output_hook <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
  output_hook(gsub("[ \\t]+\\n", "\n", x), options)
})
library(freqtables)
library(dplyr)
```

# freqtables

Quick, composable frequency tables for categorical variables in dplyr pipelines.
Use `freq_tbl()` for counts and proportions, `wald_ci()` or `logit_ci()`
to add intervals, and `freq_table()` to do both. Use `freq_xtab()` for
contingency count matrices and grouped arrays.

This checkout contains the **0.2.0.9000 development API**. It changes grouping
syntax and output defaults from 0.1.1. The code shown here applies to this
development version; the currently released version may have the previous API.

## Installation

Install the released version with:

```r
install.packages("freqtables")
```

To use this checkout's development API, build and install the local package
after its dependencies are available, or install the development branch you
intend to use. These changes are prepared locally and have not been published.

## Counts and Intervals

```{r counts}
freq_tbl(mtcars, am)
mtcars |> group_by(cyl) |> freq_tbl(am)
mtcars |> group_by(cyl) |> freq_table(am)
```

One outcome is supplied to the function; `group_by()` provides any grouping
variables. Default results use proportions. Grouped tables include cell counts
(`n`), group totals (`n_group`), and within-group proportions
(`prop_group`). Overall cell proportions are optional.

```{r options}
mtcars |> group_by(cyl, vs) |>
  freq_table(am, percent = TRUE, overall = TRUE)
freq_tbl(mtcars, am) |> wald_ci(percent_ci = 99)
```

The t-based interval convention is preserved. Each group's standard error and
critical value now both use its own sample size. See the interval vignette for
boundary behavior and formulas.

## Contingency Matrices and Arrays

`freq_xtab()` accepts two columns: exposure in rows and outcome in columns.
By default it adds row and column totals. Groups become additional array
dimensions, with common categories and zero counts for absent combinations.

```{r contingency}
freq_xtab(mtcars, vs, am)
mtcars |> group_by(cyl) |> freq_xtab(vs, am, margins = FALSE)
```

Use `drop = TRUE` to remove globally unused factor levels. Missing values
count as categories under either setting. Factors control category order.
See [Contingency Tables](vignettes/contingency-tables.Rmd) for multiple groups,
subsetting, labels, empty data, and each argument.

## Tests and Formatting

```{r report}
mtcars |> group_by(am) |>
  freq_table(cyl, percent = TRUE) |>
  freq_test() |>
  freq_format("n (percent_group%)", name = "estimate", digits = 1) |>
  select(am, cyl, estimate, p_chi2_pearson)

get_group_n(mtcars, cyl == 4)
```

Filter missing observations explicitly before counting when they should not
contribute to denominators. Generic headers are available with
`generic_col_names = TRUE`; use them to stack completed analyses.

## Learn Every Function and Argument

Build/install the package with vignettes, then open:

- `vignette("descriptive_analysis", package = "freqtables")`: inputs, groups,
  missing values, factors, and every table-constructor option.
- `vignette("confidence-intervals", package = "freqtables")`: both helpers,
  formulas, scales, confidence levels, and boundary cases.
- `vignette("using_freq_test", package = "freqtables")`: one-way/two-way
  methods, assumptions, unsupported arguments, and complete pipelines.
- `vignette("formatting", package = "freqtables")`: recipes, decimal places,
  column names, and all sample-size filter options.
- `vignette("migration", package = "freqtables")`: old/new syntax and columns.
- `vignette("contingency-tables", package = "freqtables")`: count matrices,
  grouped arrays, margins, ordering, missing values, and unused levels.

Source files are in [vignettes/](vignettes/). The exact function/argument mapping
is [inst/vignette-coverage.csv](inst/vignette-coverage.csv); development checks
compare it with live signatures. API decisions are recorded in
[dev/api-specification.md](dev/api-specification.md).

## Migration from 0.1.1

Replace `freq_table(mtcars, cyl, am)` with
`mtcars |> group_by(cyl) |> freq_table(am)`. Old multiple-variable calls
now error with this guidance. Options must be named. Use `percent = TRUE`
for percentage-scale output, and request `overall`, `se`, and
`critical_value` when needed. A rich generic table is available without
changing the compact defaults:

```{r migration}
mtcars |> group_by(cyl) |>
  freq_table(am, percent = TRUE, overall = TRUE, se = TRUE,
             critical_value = TRUE, generic_col_names = TRUE)
```
