---
title: "wiki_development_notes"
---

# Overview

- Working my way through the Wiki development notes (https://github.com/brad-cannell/freqtables/wiki/Development-Notes). 

- At this point, don't worry about great documentation. Just get things working. You can document them and add them to R scripts later.

- Having said that, do try to write modularized code -- break the main function up into smaller functions that can individually be maintained and tested.

- Delete this before submitting to CRAN.

# Load packages

Don't load packages. Use the double colon syntax like you would in an R script. 

# Load data 

For testing

```{r}
data(freq_study)
```

# freq_tbl

Create the freq_tbl function.

* We may want to come back and add a .col_headers = "col_names" argument that users can adjust if they want generic column headers instead.

* We can add an overall = FALSE argument that users who want an overall percentage can adjust.

```{r}
freq_tbl <- function(.data, .freq_var, .drop = FALSE, .percent = FALSE) {
  
  # ===========================================================================
  # Prevents R CMD check: "no visible binding for global variable ‘.’"
  # ===========================================================================
  # Fill this in later.
  
  # ===========================================================================
  # Setup for tidy evaluation
  # ===========================================================================
  .freq_var <- rlang::enquo(.freq_var)
  
  # ===========================================================================
  # Data Checks
  # ===========================================================================
  
  # ---------------------------------------------------------------------------
  # Check to see what type of object is being passed to .data
  #   - Should be a data frame.
  #   - For example, some people use "attach". That won't work with freq_tbl()
  # ---------------------------------------------------------------------------
  .data_class <- class(.data)
  if (!("data.frame" %in% .data_class)) {
    stop(
      paste0(
        "freq_tbl expects the object passed to the .data argument (the",
        " first argument) to be a data frame. Currently, the object being",
        " passed to .data has the class: ", .data_class, ". Please use the",
        " form mtcars |> freq_table(am) or freq_table(mtcars, am)."
      )
    )
  }
  
  # ---------------------------------------------------------------------------
  # Check to make sure there was a value passed to the .freq_var argument.
  #   - Not necessary. If it is missing then 'argument ".freq_var" is missing, 
  #     with no default' is thrown automatically.
  # ---------------------------------------------------------------------------
  
  # ---------------------------------------------------------------------------
  # Check to make sure that .freq_var is a column that exists in .data.
  # ---------------------------------------------------------------------------
  .freq_var_chr <- rlang::as_name(.freq_var)
  col_names <- names(.data)
  if (!(.freq_var_chr %in% col_names)) {
    stop(
      paste0(
        "freq_tbl expects the object passed to the .freq_var argument (the",
        " second argument) to be a column of the data frame passed to .data.", 
        " The object being passed to .freq_var is called '", .freq_var_chr, 
        "' and is not a column in the data frame being passed to .data."
      )
    )
  }
  
  # ===========================================================================
  # One-way frequency table
  # ===========================================================================
  
  # Get n overall
  n_overall <- nrow(.data)
  
  # Get counts and proportions
  out <- .data |>
    dplyr::count({{ .freq_var }}, .drop = .drop) |>
    # Add proportions
    dplyr::mutate(prop = n / n_overall)

  # Convert proportion to percentage (optional)
  if (.percent) {
    out <- dplyr::mutate(out, percent = prop * 100, prop = NULL)
  }
  
  # ===========================================================================
  # Two-way frequency table
  # ===========================================================================

  # Add n_group and percent_group
  if ("grouped_df" %in% class(.data)) {
    out <- out |>
      dplyr::mutate(
        n_group    = sum(n),
        prop_group = n / sum(n),
        
      )

    # Convert group proportion to percentage (optional)
    # Grouped tibbles only
    if (.percent) {
      out <- out |>
        dplyr::mutate(
          percent_group = prop_group * 100,
          prop_group    = NULL
        )
    }
  }

  # Return tibble of results
  out
}
```

## Test freq_tbl

```{r}
# Not a data frame
# freq_study$sex |> freq_tbl() # Return informative error

# Nothing passed to .freq_var
# freq_study |> freq_tbl() # Return informative error

# A column name that doesn't exist in .data is passed to .freq_var
# freq_study |> freq_tbl(age) # Return informative error

# One-way table
# freq_study |> freq_tbl(sex)

# One-way table with .drop = TRUE
# freq_study |> freq_tbl(bmi_4cat, .drop = FALSE)
# freq_study |> freq_tbl(bmi_4cat, .drop = TRUE)

# One-way table with percentages instead of proportions
# freq_study |> freq_tbl(sex, .percent = TRUE)

# Two-way table
freq_study |> dplyr::group_by(exposure) |> freq_tbl(sex)

# Two-way table with .drop = TRUE
# freq_study |> freq_tbl(bmi_4cat, .drop = FALSE)
# freq_study |> freq_tbl(bmi_4cat, .drop = TRUE)

# Return proportion instead of percent
# freq_study |> group_by(exposure) |> freq_tbl(sex, percent = FALSE)

# Return NA as a category
# freq_study |> group_by(exposure, outcome) |> freq_tbl(sex, .drop = FALSE)

# Passing more than one column
# freq_table(freq_study, exposure, sex)
```

- When you get this working, try breaking the function up into smaller functions.


















# Dumping ground

Stuff I think I don't need anymore, but I'm not 100% sure.

## freq_tbl_data_check

Make this a separate function for easier maintenance and testing. 

```{r}
freq_tbl_data_check <- function(.data, .freq_var) {
  # ===========================================================================
  # Prevents R CMD check: "no visible binding for global variable ‘.’"
  # ===========================================================================
  # Fill this in later.
  
  # ===========================================================================
  # Check to see what type of object is being passed to .data
  #   - Should be a data frame.
  #   - For example, some people use "attach". That won't work with freq_tbl()
  # ===========================================================================
  .data_class <- class(.data)
  if (!("data.frame" %in% .data_class)) {
    stop(
      paste0(
        "freq_tbl expects the object passed to the .data argument (the",
        " first argument) to be a data frame. Currently, the object being",
        " passed to .data has the class: ", .data_class, ". Please use the",
        " form mtcars |> freq_table(am) or freq_table(mtcars, am)."
      )
    )
  }
  
  # ===========================================================================
  # Check to make sure there was a value passed to the .freq_var argument.
  # And that the value is a column that exists in .data.
  # ===========================================================================
  if (is.null(.freq_var)) {
    stop(
      "It's empty."
    )
  }
  # Defuse the user expression in .freq_var
  .freq_var <- rlang::enquo(.freq_var)
  # out <- dplyr::pull(.data, rlang::enquo(.freq_var))
  return(.freq_var)
  
  # Pass .data along if it clears data checks
  .data
}
```

## Test freq_tbl_data_check

```{r}
# Not a data frame
# freq_study$sex |> freq_tbl_data_check() # Return informative error

# Nothing passed to .freq_var
# freq_study |> freq_tbl_data_check() # Return informative error
```


Actually, I'm not sure I want to do it this way. Maybe I should put the data checks inside the main function. 
- I want to check all of the arguments.
- I don't want to have to enquo arguments multiple times.
- Still break the stats functions into multiple functions when that makes sense.
