---
title: "wiki_development_notes"
---

# Overview

- Working my way through the Wiki development notes (https://github.com/brad-cannell/freqtables/wiki/Development-Notes). 

- At this point, don't worry about great documentation. Just get things working. You can document them and add them to R scripts later.

- Having said that, do try to write modularized code -- break the main function up into smaller functions that can individually be maintained and tested.

- Delete this before submitting to CRAN.

## Process

- As the functions get longer, it's actually easier to to keep them in a separate R script and then source them than it is to scroll up and down running them and testing them.

- Start by making the functions you write accomplish the main task they are supposed to accomplish. Then, come back and put in data checks.

# Load packages

Don't load packages. Use the double colon syntax like you would in an R script. 

# Load data 

For testing

```{r}
data(freq_study)
```

# freq_tbl

Create the freq_tbl function.

- 🛑 When you get this working, try breaking the function up into smaller functions.
  - I tried playing with this. I don't think there's enough code there for this to make sense.

- Need to create an informative error when more than one column name is passed to freq_tbl().
  - Currently, it just says "object 'sex' not found".

- Need to create an informative error when nothing is passed to freq_tbl()
  - Currently, it just says "Error in as_string(x) : argument "x" is missing, with no default"

- We may want to come back and add a .col_headers = "col_names" argument that users can adjust if they want generic column headers instead.

- We can add an overall = FALSE argument that users who want an overall percentage can adjust (two-way and n-way tables).
  - Maybe. Now, I think the default should be: n, prop, n_group, prop_group
  - Annotate this in the wiki.
  
- Finish creating unit tests: tests/testthat/test-freq_tbl.R. You can use test-freq_table.R for inspiration.

```{r}
# Load functions
devtools::load_all()
test_results <- freq_tbl_test()
test_results
```

```{r}
# - Need to create an informative error when more than one column name is passed to freq_tbl().
#   - Currently, it just says "object 'sex' not found".
devtools::load_all()
freq_study |> freq_tbl(sex)
```






















# Dumping ground

Stuff I think I don't need anymore, but I'm not 100% sure.

## freq_tbl_data_check

Make this a separate function for easier maintenance and testing. 

```{r}
freq_tbl_data_check <- function(.data, .freq_var) {
  # ===========================================================================
  # Prevents R CMD check: "no visible binding for global variable ‘.’"
  # ===========================================================================
  # Fill this in later.
  
  # ===========================================================================
  # Check to see what type of object is being passed to .data
  #   - Should be a data frame.
  #   - For example, some people use "attach". That won't work with freq_tbl()
  # ===========================================================================
  .data_class <- class(.data)
  if (!("data.frame" %in% .data_class)) {
    stop(
      paste0(
        "freq_tbl expects the object passed to the .data argument (the",
        " first argument) to be a data frame. Currently, the object being",
        " passed to .data has the class: ", .data_class, ". Please use the",
        " form mtcars |> freq_table(am) or freq_table(mtcars, am)."
      )
    )
  }
  
  # ===========================================================================
  # Check to make sure there was a value passed to the .freq_var argument.
  # And that the value is a column that exists in .data.
  # ===========================================================================
  if (is.null(.freq_var)) {
    stop(
      "It's empty."
    )
  }
  # Defuse the user expression in .freq_var
  .freq_var <- rlang::enquo(.freq_var)
  # out <- dplyr::pull(.data, rlang::enquo(.freq_var))
  return(.freq_var)
  
  # Pass .data along if it clears data checks
  .data
}
```

## Test freq_tbl_data_check

```{r}
# Not a data frame
# freq_study$sex |> freq_tbl_data_check() # Return informative error

# Nothing passed to .freq_var
# freq_study |> freq_tbl_data_check() # Return informative error
```


Actually, I'm not sure I want to do it this way. Maybe I should put the data checks inside the main function. 
- I want to check all of the arguments.
- I don't want to have to enquo arguments multiple times.
- Still break the stats functions into multiple functions when that makes sense.
