---
title: "Composable Confidence Intervals"
author: "Brad Cannell"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Composable Confidence Intervals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(freqtables)
library(dplyr)
```


The interval helpers allow a pipeline to stop after counts, or add intervals
later. Both accept the same arguments: `.data`, `percent_ci`, `se`,
and `critical_value`. They share their calculations with `freq_table()`.

## Inputs and Defaults {#inputs}

`.data` is normally a complete or filtered result from `freq_tbl()`.
Its metadata retains the original sample size, so filtering a result does
not change a denominator. Grouped tables also carry explicit `n_group`.

```{r inputs}
counts <- freq_tbl(mtcars, am)
counts |> wald_ci()
counts |> logit_ci()
counts |> filter(am == 0) |> logit_ci()
```

Default output appends `lcl`/`ucl` for ungrouped results and
`lcl_group`/`ucl_group` for grouped results. If overall proportions
are present, the helpers also compute `lcl_total`/`ucl_total`.

A plain data frame is supported with explicit denominators. The required
columns are `n`, `n_total`, and `prop` (or `percent`), or
`n`, `n_group`, and `prop_group` (or `percent_group`).
Counts and denominators must be finite nonnegative integers, with
`n <= denominator`. Displayed proportions must agree with them.

```{r plain-input}
plain <- data.frame(n = c(19L, 13L), n_total = 32L, prop = c(19, 13) / 32)
wald_ci(plain)
logit_ci(plain)
plain_group <- data.frame(n = c(3L, 7L), n_group = 10L, prop_group = c(.3, .7))
wald_ci(plain_group)
logit_ci(plain_group)
```

The helpers never infer an absent original denominator by summing a
potentially filtered count table. If metadata has been removed, provide
the original `n_total` explicitly.

## Formula and Confidence Level {#level}

For continuity, both methods use the existing package convention:

```
p = n / N
SE = sqrt(p * (1 - p) / (N - 1))
t = qt((1 + percent_ci / 100) / 2, df = N - 1)
```

For a grouped interval, **N is the group size in both formulas**.
For an overall interval, it is the total number of input observations.
The former implementation used total-sample degrees of freedom for
within-group limits; those limits change with this correction.

`wald_ci()` returns `p - t * SE` and `p + t * SE`.
`logit_ci()` applies `t * SE / (p * (1 - p))` on the log-odds scale,
then transforms back with `plogis()`. This is a t-based compatibility
convention; it is not the usual normal-z binomial Wald formula.

`percent_ci = 95` is the default; use a number strictly between 0 and
100 to change it. Larger levels produce wider intervals for interior
proportions.

```{r level}
counts |> wald_ci(percent_ci = 90)
counts |> wald_ci(percent_ci = 99)
counts |> logit_ci(percent_ci = 90)
counts |> logit_ci(percent_ci = 99)

# Independently verify the first logit lower bound.
p <- 19 / 32
expected <- plogis(qlogis(p) - qt(.975, 31) *
                    sqrt(p * (1 - p) / 31) / (p * (1 - p)))
stopifnot(isTRUE(all.equal(logit_ci(counts)$lcl[1], expected)))
```

## Standard Errors and Critical Values {#details}

`se = TRUE` appends `se`, `se_group`, and/or `se_total`.
`critical_value = TRUE` appends `t_crit`, `t_crit_group`, and/or
`t_crit_total`. Each option defaults to `FALSE`; either can be enabled
without the other.

```{r details}
counts |> wald_ci(se = TRUE)
counts |> wald_ci(critical_value = TRUE)
counts |> logit_ci(se = TRUE)
counts |> logit_ci(critical_value = TRUE)
grouped <- mtcars |> group_by(cyl) |> freq_tbl(am, overall = TRUE)
grouped |> wald_ci(se = TRUE, critical_value = TRUE)
grouped |> logit_ci(se = TRUE, critical_value = TRUE)
```

## Percentage Scale and Composition {#scale}

When counts use `percent = TRUE`, bounds and standard errors also use
the 0--100 scale. Critical values are unchanged. The helpers obtain the
scale from the `prop*` or `percent*` columns; `percent_ci` always
specifies the confidence level, not the display scale.

```{r scale}
pct <- mtcars |> group_by(cyl) |> freq_tbl(am, percent = TRUE, overall = TRUE)
pct |> wald_ci(se = TRUE, critical_value = TRUE)
pct |> logit_ci(se = TRUE, critical_value = TRUE)

# Replacing Wald intervals with logit intervals does not multiply by 100 again.
pct |> wald_ci(se = TRUE) |> logit_ci()
```

Repeated calls replace interval columns, and unrequested `se*` and
`t_crit*` columns are removed. Renaming proportions or removing denominator
columns/metadata can make the table unsuitable for subsequent calculations.
Use formatting only after all computations.

## Boundaries and Errors {#boundaries}

At p = 0 or 1, logit limits are missing because the logit is infinite;
no continuity correction is applied. Wald limits equal p when N > 1.
For interior values, Wald intervals can extend beyond [0, 1] and are not
clipped. N = 0 gives an undefined proportion. N <= 1 gives missing
standard errors, critical values, and limits in either method.

```{r boundaries}
boundary <- data.frame(y = factor(rep("yes", 4), levels = c("no", "yes")))
freq_tbl(boundary, y) |> logit_ci()
freq_tbl(boundary, y) |> wald_ci()
freq_tbl(boundary[1, , drop = FALSE], y) |> wald_ci(se = TRUE)
freq_tbl(boundary[FALSE, , drop = FALSE], y) |> logit_ci()
wald_ci(data.frame(n = 1, n_total = 3, prop = 1/3))
```

Missing logit bounds do not mean the population proportion is known to be
exactly zero or one. These approximations need care with small or extreme
counts; the package does not automatically select another interval method.

```{r bad-level, error=TRUE}
counts |> logit_ci(percent_ci = 100)
```
```{r bad-count, error=TRUE}
wald_ci(data.frame(n = 6, n_total = 5, prop = 1.2))
```
```{r missing-denominator, error=TRUE}
logit_ci(data.frame(n = 19, prop = 19/32))
```

See R's [Student t distribution documentation](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/TDist.html)
for the critical-value function used here.
