---
title: "Contingency Tables with freq_xtab"
author: "Brad Cannell"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Contingency Tables with freq_xtab}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(freqtables)
library(dplyr)
```

Use `freq_xtab()` to arrange observation counts in a contingency matrix.
Exposure categories form the rows and outcome categories form the columns.
Grouping variables create additional array dimensions so each group has its
own matrix. The function counts input rows; it does not interpret an input
column named `n` as weights or accept an already aggregated count table.

## Inputs and the Source Example {#inputs}

The `.data` argument contains individual observations. Supply two distinct
columns as `exposure` and `outcome`. Here the four cells contain 3, 2, 3, and
2 observations. Factors explicitly place `Yes` before `No` on each axis.

```{r basic}
study <- data.frame(
  exposure = factor(rep(c("Yes", "Yes", "No", "No"), c(3, 2, 3, 2)),
                    levels = c("Yes", "No")),
  outcome = factor(rep(c("Yes", "No", "Yes", "No"), c(3, 2, 3, 2)),
                   levels = c("Yes", "No"))
)
x <- study |> freq_xtab(exposure, outcome)
x
stopifnot(identical(unname(x), matrix(c(3, 2, 5, 3, 2, 5, 6, 4, 10),
                                    nrow = 3, byrow = TRUE)))
```

The row totals are 5 and 5, the column totals are 6 and 4, and the grand total
is 10. The result is a numeric matrix, with column names retained as dimension
names. It is not limited to binary categories.

Literal strings and tidy-evaluation forwarding are supported too:

```{r programming}
freq_xtab(study, exposure = "exposure", outcome = "outcome")
cross_counts <- function(data, e, o) freq_xtab(data, {{ e }}, {{ o }})
cross_counts(study, exposure, outcome)
```

## Margins and Further Analysis {#margins}

`margins = TRUE` appends a `Sum` row and column by default. Set
`margins = FALSE` when you need the cell counts alone. Including totals as
additional observations would change subsequent calculations.

```{r margins}
freq_xtab(study, exposure, outcome, margins = TRUE)
cells <- freq_xtab(study, exposure, outcome, margins = FALSE)
cells
stopifnot(sum(cells) == nrow(study), identical(cells, x[1:2, 1:2]))
```

If a category is already named `Sum`, the total receives the first available
label `Sum.1`, `Sum.2`, and so on, separately on each axis. Category labels
themselves remain unchanged. The total is always the last row or column.

```{r total-labels}
labels <- data.frame(e = c("Sum", "Sum.1", "other"), o = c("a", "a", "b"))
freq_xtab(labels, e, o)
```

`freq_test()` currently accepts the tibble results from `freq_table()`;
it does not accept `freq_xtab()` matrices. Likewise, the package's existing
interval and formatting helpers target frequency-table columns. This function
provides counts; epidemiologic measures and new presentation formats remain
separate work.

## One or Multiple Groups {#grouping}

`group_by()` supplies groups. One grouping variable produces a three-dimensional
array: exposure, outcome, then group. Every group shares the same categories.

```{r one-group}
study$site <- rep(c("Site A", "Site B"), each = 5)
by_site <- study |> group_by(site) |> freq_xtab(exposure, outcome)
by_site
by_site[, , "Site A"]
stopifnot(identical(dim(by_site), c(3L, 3L, 2L)),
          all(by_site["No", , "Site A"] == 0),
          all(by_site["Yes", , "Site B"] == 0))
```

Each additional grouping variable adds a dimension, in `group_by()` order.
Here, only Site A/female and Site B/male occur. The other group combinations
have zero-filled matrices. Margins are calculated within each matrix; grouping
dimensions do not receive a `Sum` category.

```{r multiple-groups}
study$sex <- rep(c("female", "male"), each = 5)
grouped <- study |> group_by(site, sex)
by_group <- grouped |> freq_xtab(exposure, outcome)
dim(by_group)
names(dimnames(by_group))
by_group[, , "Site A", "female"]
by_group[, , "Site B", "female"]
stopifnot(identical(dim(by_group), c(3L, 3L, 2L, 2L)),
          all(by_group[, , "Site B", "female"] == 0),
          sum(by_group["Sum", "Sum", , ]) == nrow(study))
```

The returned array retains singleton dimensions. R's own subsetting rules can
drop them, so use `by_group[, , "Site A", "female", drop = FALSE]` to retain
the full array shape. This subsetting option is separate from the constructor's
`drop` option below.

## Missing Values, Ordering, and Unused Levels {#categories}

Factors use their specified level order. Other categories are sorted: numeric
values numerically, logical values `FALSE` then `TRUE`, and character values
in C-locale order. Missing values appear last, before any margin. Use factors
to request a particular ordering; `Yes` is not automatically placed first.

`drop = FALSE` retains unused factor levels on exposure, outcome, and group
axes. `drop = TRUE` removes globally unused levels, overriding the input's
grouping drop setting. Categories observed elsewhere in the input remain in
every group's matrix, even when their count in that group is zero. Both settings
represent every combination of the retained axis categories.

```{r drop}
with_missing <- data.frame(
  e = factor(c("yes", "no", NA), levels = c("yes", "no", "unused")),
  o = factor(c("yes", NA, "no"), levels = c("yes", "no", "unused")),
  g = factor(c("A", "B", NA), levels = c("A", "B", "empty"))
)
kept <- with_missing |> group_by(g) |> freq_xtab(e, o, drop = FALSE)
kept[, , "empty"]
pruned <- with_missing |> group_by(g, .drop = FALSE) |> freq_xtab(e, o, drop = TRUE)
dim(kept)
dim(pruned)
stopifnot(all(kept[, , "empty"] == 0),
          identical(dim(kept), c(5L, 5L, 4L)),
          identical(dim(pruned), c(4L, 4L, 3L)))
```

Missing exposure, outcome, and group values are counted as categories under
either `drop` setting. Numeric `NaN` and `NA` share the missing category.
Exclude observations explicitly with `filter()` when appropriate:

```{r missing}
with_missing |>
  filter(!is.na(e), !is.na(o), !is.na(g)) |>
  group_by(g) |> freq_xtab(e, o, drop = TRUE)
```

Dimension labels are character representations of values. A missing label is
an actual `NA_character_`, distinct from a literal `"NA"` or `"<NA>"` category.
Use positional indexing to select a missing category unambiguously:

```{r missing-labels}
missing_group <- which(is.na(dimnames(pruned)$g))
pruned[, , missing_group]
```

Empty data retain factor levels with zero counts when `drop = FALSE`.
Empty nonfactor axes have length zero before margins. Empty ungrouped
character data therefore return a 1-by-1 zero matrix with margins, or a
0-by-0 matrix without margins. An empty grouping axis yields no matrix slices.

```{r empty}
empty <- data.frame(e = character(), o = character())
freq_xtab(empty, e, o)
dim(freq_xtab(empty, e, o, margins = FALSE))
```

## Invalid Calls {#invalid}

`...` must be empty and options must be named. A third column or unsupported
option produces an informative error:

```{r invalid, error=TRUE}
freq_xtab(study, exposure, outcome, site)
```

```{r invalid-option, error=TRUE}
freq_xtab(study, exposure, outcome, na.rm = TRUE)
```

Both analysis columns must exist, be distinct, and not be grouping variables.
Create derived columns with `mutate()` first and ungroup rowwise data.
List, matrix, raw, and complex columns are unsupported. If distinct values
would have identical text labels (for example, nearly equal floating-point
values), recode them into categories with unique labels first. Dense arrays
grow with the product of category counts; reduce the grouping variables or
recode categories when that shape would be too large.
