---
title: "Counts, Proportions, and Grouped Frequency Tables"
author: "Brad Cannell"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Counts, Proportions, and Grouped Frequency Tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(freqtables)
library(dplyr)
```


Use `freq_tbl()` for counts and proportions and `freq_table()` when you also
want confidence intervals. Both analyze one outcome; preceding `group_by()`
calls specify any grouping variables. All examples use local data frames.

## Inputs and Default Output {#inputs}

Here are 16 observations whose counts can be checked directly: group A has
3 "no" and 7 "yes"; group B has 4 "no" and 2 "yes". The overall counts are
7 "no" and 9 "yes". The `.data` argument is the data frame (supplied by the
pipe below), and `.freq_var` is its outcome column.

```{r basic}
study <- data.frame(
  group = rep(c("A", "A", "B", "B"), c(3, 7, 4, 2)),
  outcome = rep(c("no", "yes", "no", "yes"), c(3, 7, 4, 2))
)
freq_tbl(study, outcome)
study |> freq_table(outcome)
```

The first result contains `outcome`, `n`, and `prop`: 7/16 and 9/16.
The second adds `lcl` and `ucl`. Proportions and interval limits are
numbers on the 0--1 scale. Rounding in printed tibbles does not round stored
values. Outcomes retain their original type, including factor levels.

The column may also be a literal string. To forward an unquoted column
through your own function, embrace it with `{{ }}`:

```{r programming}
freq_tbl(study, "outcome")
freq_table(study, "outcome")
count_outcome <- function(data, variable) freq_tbl(data, {{ variable }})
interval_outcome <- function(data, variable) freq_table(data, {{ variable }})
count_outcome(study, outcome)
interval_outcome(study, outcome)
```

## One or Multiple Grouping Variables {#grouping}

```{r grouping}
grouped <- study |> group_by(group)
grouped |> freq_tbl(outcome)
grouped |> freq_table(outcome)
```

`n` is the cell count; `n_group` is 10 for A and 6 for B. Thus the
within-group proportions are 3/10, 7/10, 4/6, and 2/6. Each group's
proportions sum to 1. `freq_table()` adds `lcl_group` and `ucl_group`.

The same rule applies to each combination of multiple groups:

```{r multiple-groups}
study$band <- rep(c("young", "old"), 8)
study |> group_by(group, band) |> freq_tbl(outcome)
study |> group_by(group, band) |> freq_table(outcome)
```

Results are ungrouped tibbles. This avoids carrying implicit grouping into
later calculations. To summarize the results by group, call `group_by()`
again. A one-person group can have a defined proportion but has missing
intervals under the package's t-based convention.

## Percentages and Overall Results {#display}

`percent = FALSE` is the default in both functions. Set it to `TRUE`
to multiply proportions, interval bounds, and requested standard errors by
100. Count columns and t critical values do not change. Proportion column
names switch from `prop*` to `percent*`.

```{r percentages}
grouped |> freq_tbl(outcome, percent = TRUE)
grouped |> freq_table(outcome, percent = TRUE)
freq_tbl(study, outcome, percent = TRUE)
freq_table(study, outcome, percent = TRUE)
```

`overall = FALSE` keeps grouped output focused on within-group results.
With `TRUE`, `n_total` is the total input rows, and `prop_total` is
the **cell count divided by that total**. For A/"no", this is 3/16, not
the marginal 7/16 for all "no" outcomes. `freq_table()` also adds
overall interval bounds. For ungrouped data, `overall = TRUE` adds only
`n_total`, because `prop` and its intervals are already overall.

```{r overall}
grouped |> freq_tbl(outcome, overall = TRUE)
grouped |> freq_table(outcome, overall = TRUE)
freq_tbl(study, outcome, overall = TRUE)
freq_table(study, outcome, overall = TRUE)
grouped |> freq_table(outcome, overall = TRUE, percent = TRUE)
```

## Missing Values and Unobserved Factor Levels {#missing}

Missing outcomes and grouping values count as categories by default. To
exclude them, filter the original observations. Neither function has an
`na.rm` argument. The denominators always describe the rows after filtering.

```{r missing}
with_missing <- data.frame(
  group = factor(c("A", "A", NA), levels = c("A", "B")),
  outcome = factor(c("no", NA, "yes"), levels = c("no", "yes", "unused"))
)
with_missing |> group_by(group) |> freq_tbl(outcome)
with_missing |> group_by(group) |> freq_table(outcome)
with_missing |>
  filter(!is.na(group), !is.na(outcome)) |>
  group_by(group) |> freq_tbl(outcome)
```

`drop = FALSE` retains unobserved factor levels and empty factor-group
combinations with `n = 0`, following
[dplyr's count semantics](https://dplyr.tidyverse.org/reference/count.html).
A completely empty group has `n_group = 0` and missing proportions and
intervals. This option is independent of missing-value removal.

```{r drop}
freq_tbl(with_missing, outcome, drop = FALSE)
freq_tbl(with_missing, outcome, drop = TRUE)
freq_table(with_missing, outcome, drop = FALSE)
freq_table(with_missing, outcome, drop = TRUE)
with_missing |> group_by(group, .drop = FALSE) |>
  freq_table(outcome, drop = TRUE)
```

An empty data frame with a factor outcome can therefore produce explicit
zero rows; an empty character outcome produces a zero-row result.

## Generic Headers for Programming {#generic}

`generic_col_names = FALSE` uses variable names as headers and preserves
their types. With `TRUE`, the outcome becomes `col`/`cat`, and
groups become `group_01_col`/`group_01_cat`, `group_02_col`/
`group_02_cat`, and so on in `group_by()` order. All category columns
become character, allowing numeric and character outcomes to be stacked.

```{r generic}
grouped |> freq_tbl(outcome, generic_col_names = TRUE)
grouped |> freq_table(outcome, generic_col_names = TRUE)
study |> group_by(group, band) |>
  freq_table(outcome, generic_col_names = TRUE)
bind_rows(
  freq_tbl(study, outcome, generic_col_names = TRUE),
  freq_tbl(mtcars, am, generic_col_names = TRUE)
)
```

Calculate intervals and tests **before** stacking separate analyses. Metadata
belongs to one analysis and is not a registry for a stack of tables.

## Confidence-Interval Controls {#interval-controls}

Only `freq_table()` has interval controls. `percent_ci = 95` means
95% confidence, regardless of `percent`. Any numeric level strictly
between 0 and 100 is accepted. `ci_type` is either `"logit"`
(default) or `"wald"`; both work with any number of groups.

```{r interval-controls}
grouped |> freq_table(outcome, percent_ci = 90, ci_type = "logit")
grouped |> freq_table(outcome, percent_ci = 99, ci_type = "wald")
```

`se = FALSE` and `critical_value = FALSE` omit intermediate
statistics. Enable them independently to include standard errors and
Student-t critical values. These are also available for overall results:

```{r details}
grouped |> freq_table(outcome, se = TRUE)
grouped |> freq_table(outcome, critical_value = TRUE)
grouped |> freq_table(outcome, se = TRUE, critical_value = TRUE,
                       overall = TRUE, percent = TRUE)
```

For formulas and boundary behavior, read
[Confidence Intervals](confidence-intervals.html).

## Invalid Calls and Reserved Arguments {#invalid}

`...` must be empty: both functions accept one outcome, with all other
options supplied by name. An old two-variable call gives a migration error.
Incorrectly named options are errors too.

```{r invalid, error=TRUE}
freq_tbl(study, group, outcome)
```
```{r invalid-table, error=TRUE}
freq_table(study, group, outcome)
```
```{r invalid-option, error=TRUE}
freq_table(study, outcome, na.rm = TRUE)
```

Create derived variables with `mutate()` first. Do not use the outcome as
its own grouping variable. Rowwise input requires `ungroup()`. Analysis
names that conflict with generated columns (for example `n` or `prop`)
must be renamed first, so result columns never silently overwrite categories.
