---
title: "Formatting Results and Sample Sizes"
author: "Brad Cannell"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Formatting Results and Sample Sizes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(freqtables)
library(dplyr)
```


Compute counts, intervals, and tests before preparing display strings.
`freq_format()` appends presentation text; `get_group_n()` creates
sample-size labels from observation-level data.

## Formatting Input and Recipes {#recipes}

`.data` is a data frame, usually a result from `freq_tbl()`,
`freq_table()`, or `freq_test()`. `recipe` is one quoted string
mixing column names with punctuation or other literal text.

```{r recipes}
estimates <- freq_table(mtcars, am, percent = TRUE)
estimates |> freq_format("n (percent%)")
estimates |> freq_format("percent (lcl - ucl)")
mtcars |> group_by(cyl) |> freq_table(am) |>
  freq_format("n: prop_group (lcl_group - ucl_group)")
```

The recipe must use column names that exist in the input. Unmatched words
are retained as literal text, so a misspelled name will print unchanged.
Tokens are separated at word boundaries. Rename columns containing spaces
or punctuation before putting them in recipes. Proportion and percentage
column names are intentionally different.

## Column Name and Decimal Places {#digits}

`name = NA` uses the default `formatted_stats`. Supply a nonempty
string to name the new column. If that column exists, it is replaced.

`digits = NA` uses character conversion without explicit rounding.
A nonnegative integer requests that many decimal places for numeric
double columns, including trailing zeros and thousands separators.
Integer count columns retain their integer representation.
Character categories are left as text.

```{r digits}
estimates |> freq_format("percent (lcl - ucl)", name = NA, digits = NA)
estimates |> freq_format("percent (lcl - ucl)", name = "percent_ci", digits = 2)
estimates |> freq_format("n (percent%)", digits = 0)
freq_table(mtcars, am) |> freq_format("prop (lcl - ucl)", digits = 3)
estimates |> freq_format("n", name = "label") |>
  freq_format("n (percent%)", name = "label", digits = 1)
```

The returned data frame retains all original statistics; only the new
string is rounded. To show selected columns, follow with `select()`.

```{r select-display}
estimates |> freq_format("percent (lcl - ucl)", name = "estimate", digits = 2) |>
  select(am, n, estimate)
```

Zero-row input receives an empty character column. Missing numeric limits
are displayed as missing-value text. Invalid recipes or negative decimal
places produce informative errors.

```{r format-error, error=TRUE}
estimates |> freq_format("n", digits = -1)
```

## Sample-Size Input and Filter Expressions {#sample-size}

For `get_group_n()`, `.data` is the original observation-level data.
`...` accepts zero or more `dplyr::filter()` expressions.
With no filters, all input observations are counted. Multiple expressions
are combined with AND. Missing logical filter results are excluded,
following `filter()`. The result is a character vector, not a table.

```{r sample-size}
get_group_n(mtcars)
mtcars |> get_group_n(cyl == 4)
mtcars |> get_group_n(cyl == 4, am == 1)
mtcars |> get_group_n(cyl == 99)
```

These return "N = 32", "N = 11", "N = 8", and "N = 0". Passing a
frequency table would count its category rows, not its observations.

For grouped data, one label is returned per retained group, in dplyr's
group order. Retain a companion group-label table if those labels will be
used elsewhere.

```{r grouped-labels}
mtcars |> group_by(cyl) |> get_group_n(am == 1)
data.frame(cyl = c(4, 6, 8),
           label = mtcars |> group_by(cyl) |> get_group_n(am == 1))
```

## End-to-End Reporting {#report}

```{r report}
sample_label <- mtcars |> get_group_n(!is.na(am), !is.na(cyl))
sample_label
mtcars |> filter(!is.na(am), !is.na(cyl)) |>
  group_by(am) |>
  freq_table(cyl, percent = TRUE) |>
  freq_test() |>
  freq_format("n (percent_group%)", name = "count_percent", digits = 1) |>
  select(am, cyl, count_percent, p_chi2_pearson)
```
