---
title: "Migrating to the Compact Frequency-Table API"
author: "Brad Cannell"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Migrating to the Compact Frequency-Table API}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(freqtables)
library(dplyr)
```


The development API changes defaults and grouping syntax. Update analysis
scripts and recipes together. Use this guide when moving from version
0.1.1 or the earlier experimental branches.

## Grouping and Named Options {#grouping}

An old call `freq_table(mtcars, cyl, am)` now produces an error showing
the replacement. It is not a warning followed by a calculation.

```{r migration-group}
mtcars |> group_by(cyl) |> freq_table(am)
```

The one outcome is supplied as `.freq_var`. All subsequent options must
be named, because `...` is reserved for catching old calls and typos.
For example, replace a positional confidence-level argument with
`percent_ci = 99`. The experimental `.drop` and `.percent`
names become `drop` and `percent`. There is one header option:
`generic_col_names`; proposals such as `col_headers` are not aliases.

```{r migration-options}
freq_tbl(mtcars, am, drop = TRUE, percent = TRUE)
freq_table(mtcars, am, percent_ci = 99, ci_type = "wald")
```

## Output Names and Richer Tables {#columns}

| Previous result | Current equivalent |
|---|---|
| `var` / `cat` | Outcome-named column, or `col` / `cat` with generic headers |
| `row_var` / `row_cat` | Group-named column, or `group_01_col` / `group_01_cat` |
| `col_var` / `col_cat` | Outcome-named column, or `col` / `cat` |
| `n_row` | `n_group` |
| `percent_row` | `prop_group`, or `percent_group` with `percent = TRUE` |
| `lcl_row` / `ucl_row` | `lcl_group` / `ucl_group` |
| Always-present overall/detail columns | Request `overall`, `se`, and `critical_value` |

All proportions, standard errors, and interval bounds default to the 0--1
scale. To request a rich, percentage-scale table with generic headers:

```{r rich-output}
mtcars |> group_by(cyl) |>
  freq_table(am, percent = TRUE, overall = TRUE, se = TRUE,
             critical_value = TRUE, generic_col_names = TRUE)
```

If downstream code still expects old names, rename explicitly. This is
a compatibility mapping for a two-way result:

```{r legacy-columns}
legacy <- mtcars |> group_by(cyl) |>
  freq_table(am, percent = TRUE, overall = TRUE, se = TRUE,
             critical_value = TRUE, generic_col_names = TRUE) |>
  rename(row_var = group_01_col, row_cat = group_01_cat,
         col_var = col, col_cat = cat, n_row = n_group,
         percent_row = percent_group, se_row = se_group,
         t_crit_row = t_crit_group, lcl_row = lcl_group, ucl_row = ucl_group)
legacy
```

Perform `freq_test()` **before** this renaming, because test methods need
the current category schema and metadata. Update `freq_format()`
recipes to reference the names actually in your table.

## Statistical Continuity and Deliberate Changes {#statistics}

Ungrouped interval calculations preserve the t-based `N - 1` convention.
Grouped calculations now use `n_group - 1` for both the standard-error
denominator and degrees of freedom; the previous total-sample critical
value is no longer used for a subgroup. Therefore, old grouped interval
values cannot be recovered merely by renaming columns or multiplying by 100.

Logit bounds at p = 0 or 1 are missing without a continuity correction.
Intervals for N <= 1 are missing. Wald bounds are not clipped. Explicit
`filter()` calls determine how missing observations affect denominators.

## Available Functions and Further Examples {#coverage}

The eight exported functions in this development version are:

| Function | Worked examples |
|---|---|
| `freq_xtab()` | [Contingency matrices and grouped arrays](contingency-tables.html) |
| `freq_tbl()` | [Counts and grouping](descriptive_analysis.html) |
| `freq_table()` | [Counts, intervals, and every option](descriptive_analysis.html) |
| `wald_ci()` | [Composable confidence intervals](confidence-intervals.html) |
| `logit_ci()` | [Composable confidence intervals](confidence-intervals.html) |
| `freq_test()` and its S3 methods | [Testing frequency tables](using_freq_test.html) |
| `freq_format()` | [Formatting results](formatting.html) |
| `get_group_n()` | [Sample-size labels](formatting.html#sample-size) |

A machine-checked function/argument coverage inventory is distributed at
`system.file("vignette-coverage.csv", package = "freqtables")`.
Every inventory row points to a source section and executable example
chunk. The development check compares it with actual exported signatures
and registered S3 methods.

`freq_tables()`, epidemiologic measures, and additional presentation formats
remain separate proposals. They are not exported by this development version. Existing experimental source is preserved under `dev/`
or on its original branch.
