---
title: "Using the freq_table Function"
author: "Brad Cannell"
date: "Created: 2026-03-19 <br> Updated: `r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    css: styles.css
vignette: >
  %\VignetteIndexEntry{Using the freq_table Function}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(dplyr)
library(freqtables)

data(freq_study)

analysis_df <- freq_study |>
  filter(!is.na(sex), !is.na(outcome))

multi_group_df <- freq_study |>
  filter(!is.na(age_group), !is.na(sex), !is.na(outcome))

normalize_tbl <- function(x) {
  x |>
    mutate(across(where(is.numeric), as.numeric))
}

same_tbl <- function(x, y) {
  isTRUE(
    all.equal(
      as.data.frame(normalize_tbl(x)),
      as.data.frame(normalize_tbl(y)),
      check.attributes = FALSE
    )
  )
}
```

`freq_table()` is the confidence-interval version of `freq_tbl()`. It returns counts and proportions plus lower and upper confidence limits, while still fitting naturally into a `dplyr` pipeline.

This vignette covers the full `freq_table()` interface and verifies the results with explicit `dplyr::count()` pipelines plus base R math from `stats::qt()`. In the verification chunks below, `digits = NULL` is used so the comparisons are made on full numeric precision rather than display formatting.

# Default one-way output

By default, `freq_table()` returns counts, proportions, and logit-transformed confidence intervals.

```{r}
freq_study |>
  freq_table(outcome)
```

```{r}
one_way_counts <- freq_study |>
  count(outcome, .drop = FALSE) |>
  mutate(
    n_total = sum(n),
    prop = n / n_total
  )

one_way_logit <- freq_study |>
  freq_table(outcome, digits = NULL)

one_way_logit_check <- one_way_counts |>
  mutate(
    se = sqrt(prop * (1 - prop) / (n_total - 1)),
    t_crit = stats::qt(0.975, df = n_total - 1),
    prop_log = log(prop) - log(1 - prop),
    se_log = se / (prop * (1 - prop)),
    lcl_log = prop_log - t_crit * se_log,
    ucl_log = prop_log + t_crit * se_log,
    lcl = exp(lcl_log) / (1 + exp(lcl_log)),
    ucl = exp(ucl_log) / (1 + exp(ucl_log))
  ) |>
  select(outcome, n, prop, lcl, ucl)

same_tbl(one_way_logit, one_way_logit_check)
```

# Confidence interval options

The default `ci_type` is `"logit"`, but `"wald"` is also available.

```{r}
freq_study |>
  freq_table(outcome, ci_type = "wald")
```

```{r}
one_way_wald <- freq_study |>
  freq_table(outcome, ci_type = "wald", digits = NULL)

one_way_wald_check <- one_way_counts |>
  mutate(
    se = sqrt(prop * (1 - prop) / (n_total - 1)),
    t_crit = stats::qt(0.975, df = n_total - 1),
    lcl = prop - t_crit * se,
    ucl = prop + t_crit * se
  ) |>
  select(outcome, n, prop, lcl, ucl)

same_tbl(one_way_wald, one_way_wald_check)
```

Use `percent_ci` to request a different confidence level.

```{r}
freq_study |>
  freq_table(outcome, percent_ci = 99)
```

```{r}
one_way_logit_99 <- freq_study |>
  freq_table(outcome, percent_ci = 99, digits = NULL)

one_way_logit_99_check <- one_way_counts |>
  mutate(
    se = sqrt(prop * (1 - prop) / (n_total - 1)),
    t_crit = stats::qt(0.995, df = n_total - 1),
    prop_log = log(prop) - log(1 - prop),
    se_log = se / (prop * (1 - prop)),
    lcl_log = prop_log - t_crit * se_log,
    ucl_log = prop_log + t_crit * se_log,
    lcl = exp(lcl_log) / (1 + exp(lcl_log)),
    ucl = exp(ucl_log) / (1 + exp(ucl_log))
  ) |>
  select(outcome, n, prop, lcl, ucl)

same_tbl(one_way_logit_99, one_way_logit_99_check)
```

By default, `freq_table()` hides the intermediate standard error and critical value columns. Use `se = TRUE` and `critical_value = TRUE` to keep them.

```{r}
freq_study |>
  freq_table(
    outcome,
    ci_type = "wald",
    se = TRUE,
    critical_value = TRUE
  )
```

```{r}
one_way_wald_detail <- freq_study |>
  freq_table(
    outcome,
    ci_type = "wald",
    se = TRUE,
    critical_value = TRUE,
    digits = NULL
  )

one_way_wald_detail_check <- one_way_counts |>
  mutate(
    se = sqrt(prop * (1 - prop) / (n_total - 1)),
    t_crit = stats::qt(0.975, df = n_total - 1),
    lcl = prop - t_crit * se,
    ucl = prop + t_crit * se
  ) |>
  select(outcome, n, prop, se, t_crit, lcl, ucl)

same_tbl(one_way_wald_detail, one_way_wald_detail_check)
```

# Percentages, totals, and display formatting

Use `percent = TRUE` to scale the proportion and confidence-interval columns from 0 to 100.

```{r}
freq_study |>
  freq_table(outcome, percent = TRUE)
```

Use `digits` to control display precision. The columns remain numeric, so you can still calculate with them.

```{r}
display_tbl <- freq_study |>
  freq_table(outcome, percent = TRUE, digits = 4)

display_tbl
```

```{r}
is.numeric(display_tbl$percent)
sum(display_tbl$percent)
```

Use `overall = TRUE` to add overall denominators. For one-way output that means `n_total`; for grouped output it adds `n_total` and `prop_total` or `percent_total`.

```{r}
freq_study |>
  freq_table(outcome, overall = TRUE)
```

# Grouped frequency tables

Grouped analyses are done with `group_by()`, not by passing multiple analysis variables directly to `freq_table()`.

```{r}
analysis_df |>
  group_by(sex) |>
  freq_table(outcome)
```

Within grouped output:

- `n_group` is the denominator within each group.
- `prop_group` is the within-group proportion.
- `lcl_group` and `ucl_group` are confidence limits around `prop_group`.

```{r}
grouped_counts <- analysis_df |>
  count(sex, outcome, .drop = FALSE) |>
  group_by(sex) |>
  mutate(
    n_group = sum(n),
    prop_group = n / n_group
  ) |>
  ungroup()

grouped_logit <- analysis_df |>
  group_by(sex) |>
  freq_table(outcome, digits = NULL)

grouped_logit_check <- grouped_counts |>
  mutate(
    se_group = sqrt(prop_group * (1 - prop_group) / (n_group - 1)),
    t_crit_group = stats::qt(0.975, df = n_group - 1),
    prop_group_log = log(prop_group) - log(1 - prop_group),
    se_group_log = se_group / (prop_group * (1 - prop_group)),
    lcl_group_log = prop_group_log - t_crit_group * se_group_log,
    ucl_group_log = prop_group_log + t_crit_group * se_group_log,
    lcl_group = exp(lcl_group_log) / (1 + exp(lcl_group_log)),
    ucl_group = exp(ucl_group_log) / (1 + exp(ucl_group_log))
  ) |>
  select(sex, outcome, n, n_group, prop_group, lcl_group, ucl_group)

same_tbl(grouped_logit, grouped_logit_check)
```

Use `overall = TRUE` to add overall counts and proportions to grouped output, and combine it with `percent = TRUE` to put all proportion-related columns on the percent scale. The confidence intervals stay group-level.

```{r}
analysis_df |>
  group_by(sex) |>
  freq_table(outcome, overall = TRUE, percent = TRUE)
```

```{r}
grouped_overall_percent <- analysis_df |>
  group_by(sex) |>
  freq_table(outcome, overall = TRUE, percent = TRUE, digits = NULL)

grouped_overall_percent_check <- grouped_counts |>
  mutate(
    se_group = sqrt(prop_group * (1 - prop_group) / (n_group - 1)),
    t_crit_group = stats::qt(0.975, df = n_group - 1),
    prop_group_log = log(prop_group) - log(1 - prop_group),
    se_group_log = se_group / (prop_group * (1 - prop_group)),
    lcl_group_log = prop_group_log - t_crit_group * se_group_log,
    ucl_group_log = prop_group_log + t_crit_group * se_group_log,
    lcl_group = exp(lcl_group_log) / (1 + exp(lcl_group_log)),
    ucl_group = exp(ucl_group_log) / (1 + exp(ucl_group_log))
  ) |>
  mutate(
    n_total = sum(n),
    prop_total = n / n_total,
    percent_group = 100 * prop_group,
    percent_total = 100 * prop_total,
    lcl_group = 100 * lcl_group,
    ucl_group = 100 * ucl_group
  ) |>
  select(sex, outcome, n, n_group, percent_group, n_total, percent_total, lcl_group, ucl_group)

same_tbl(grouped_overall_percent, grouped_overall_percent_check)
```

If missing values should not contribute to the denominators, filter them out before calling `freq_table()`.

```{r}
analysis_df |>
  group_by(sex) |>
  freq_table(outcome)
```

Trying to pass more than one analysis variable gives an informative error.

```{r, error = TRUE}
freq_study |>
  freq_table(sex, outcome)
```

# Generic column names

By default, `freq_table()` uses the original data column names. When you want to stack results across variables, `generic_col_names = TRUE` is often more convenient.

```{r}
freq_study |>
  freq_table(outcome, generic_col_names = TRUE)
```

```{r}
generic_table <- freq_study |>
  freq_table(outcome, generic_col_names = TRUE, digits = NULL)

generic_table_check <- one_way_logit_check |>
  mutate(
    col = "outcome",
    cat = as.character(outcome)
  ) |>
  select(col, cat, n, prop, lcl, ucl)

same_tbl(generic_table, generic_table_check)
```

One practical pattern is row-binding several one-way tables together.

```{r}
bind_rows(
  freq_study |>
    freq_table(outcome, generic_col_names = TRUE),
  freq_study |>
    freq_table(exposure, generic_col_names = TRUE)
)
```

# Dropping unused factor levels

The `drop` argument is passed through to `dplyr::count(.drop = )`.

The `bmi_4cat` variable in `freq_study` includes an unused factor level, `"Extreme Obese"`, so it is a good way to see the effect of `drop`. Here the Wald interval is easier to inspect because zero-count categories stay finite.

```{r}
freq_study |>
  freq_table(bmi_4cat, ci_type = "wald")
```

```{r}
freq_study |>
  freq_table(bmi_4cat, ci_type = "wald", drop = TRUE)
```

```{r}
bmi_wald_keep <- freq_study |>
  freq_table(bmi_4cat, ci_type = "wald", digits = NULL)

bmi_wald_keep_check <- freq_study |>
  count(bmi_4cat, .drop = FALSE) |>
  mutate(
    n_total = sum(n),
    prop = n / n_total,
    se = sqrt(prop * (1 - prop) / (n_total - 1)),
    t_crit = stats::qt(0.975, df = n_total - 1),
    lcl = prop - t_crit * se,
    ucl = prop + t_crit * se
  ) |>
  select(bmi_4cat, n, prop, lcl, ucl)

bmi_wald_drop <- freq_study |>
  freq_table(bmi_4cat, ci_type = "wald", drop = TRUE, digits = NULL)

bmi_wald_drop_check <- freq_study |>
  count(bmi_4cat, .drop = TRUE) |>
  mutate(
    n_total = sum(n),
    prop = n / n_total,
    se = sqrt(prop * (1 - prop) / (n_total - 1)),
    t_crit = stats::qt(0.975, df = n_total - 1),
    lcl = prop - t_crit * se,
    ucl = prop + t_crit * se
  ) |>
  select(bmi_4cat, n, prop, lcl, ucl)

same_tbl(bmi_wald_keep, bmi_wald_keep_check)
same_tbl(bmi_wald_drop, bmi_wald_drop_check)
```

# Multiple grouping variables

`freq_table()` also works with more than one grouping variable. In that case, `n_group`, `prop_group`, and the grouped confidence intervals are calculated within each unique combination of grouping variables.

```{r}
multi_group_df |>
  group_by(age_group, sex) |>
  freq_table(outcome)
```

```{r}
multi_group_counts <- multi_group_df |>
  count(age_group, sex, outcome, .drop = FALSE) |>
  group_by(age_group, sex) |>
  mutate(
    n_group = sum(n),
    prop_group = n / n_group
  ) |>
  ungroup()

multi_group_logit <- multi_group_df |>
  group_by(age_group, sex) |>
  freq_table(outcome, digits = NULL)

multi_group_logit_check <- multi_group_counts |>
  mutate(
    se_group = sqrt(prop_group * (1 - prop_group) / (n_group - 1)),
    t_crit_group = stats::qt(0.975, df = n_group - 1),
    prop_group_log = log(prop_group) - log(1 - prop_group),
    se_group_log = se_group / (prop_group * (1 - prop_group)),
    lcl_group_log = prop_group_log - t_crit_group * se_group_log,
    ucl_group_log = prop_group_log + t_crit_group * se_group_log,
    lcl_group = exp(lcl_group_log) / (1 + exp(lcl_group_log)),
    ucl_group = exp(ucl_group_log) / (1 + exp(ucl_group_log))
  ) |>
  select(age_group, sex, outcome, n, n_group, prop_group, lcl_group, ucl_group)

same_tbl(multi_group_logit, multi_group_logit_check)
```

# Summary

`freq_table()` is most useful when you want a compact, pipeline-friendly table of counts, proportions, and confidence intervals. The main patterns are:

- Use the default logit intervals for the standard output.
- Switch to `ci_type = "wald"` when that interval definition is the better fit for your workflow.
- Use `group_by()` to define grouped denominators.
- Use `overall = TRUE` when you also want overall counts and proportions in grouped output.
- Use `percent = TRUE` and `digits` to control presentation without losing numeric columns.
- Use `generic_col_names = TRUE` when you want to stack results across variables.
- Use `drop = TRUE` when you do not want unused factor levels retained.
