---
title: "Testing Frequency Tables"
author: "Brad Cannell"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Testing Frequency Tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(freqtables)
library(dplyr)
```


`freq_test()` is an S3 generic with methods for one-way and two-way
`freq_table()` results. The method is selected from the result class:
no grouping variable gives a goodness-of-fit test against equal category
probabilities; one grouping variable gives a test of independence.

## Input and One-Way Method {#one-way}

`.data` must be a complete result with its category/count columns and
analysis metadata intact. This applies to the generic and both supported
methods. Filter the original observations to define an analysis population,
then calculate the table and test; filtering table rows can invalidate a test.

```{r one-way}
one <- freq_table(mtcars, am)
one |> freq_test()
one |> freq_test() |> select(am, n, prop, n_expected, p_chi2_pearson)
```

The two counts are 19 and 13. Under equal proportions, each expected count
is 16. The Pearson statistic is 1.125 with 1 degree of freedom, giving
p approximately 0.289. The same overall statistic and p-value are repeated
on each row to make them available for formatting.

This tests an equal-probability null; it does not accept arbitrary expected
probabilities. The generic forwards `...` to the selected method, but
both methods currently require it to be empty. No `method` argument or
other extra option is supported.

```{r unused-dots, error=TRUE}
one |> freq_test(method = "fisher")
```

## Two-Way Method and Small Expected Counts {#two-way}

```{r two-way}
two <- mtcars |> group_by(am) |> freq_table(cyl)
tested <- two |> freq_test()
tested |> select(am, cyl, n, n_expected, p_chi2_pearson, p_fisher)
```

This tests independence between `am` and `cyl`. Output includes
`n_col` (outcome marginal counts), expected cell counts,
`chi2_contrib`, the summed statistic, numbers of rows/columns (`r`,
`c`), degrees of freedom, and `p_chi2_pearson`. The Pearson test is
uncorrected, including for a 2 by 2 table.

When any expected cell count is <= 5, the function emits a message and
adds `p_fisher` alongside the Pearson result. Fisher's calculation uses
the full rectangular contingency table. It is not restricted to two
group categories. As with R's `fisher.test()`, a large exact calculation
may fail; that error is propagated.

```{r rectangular}
mtcars |> group_by(cyl) |> freq_table(am) |> freq_test()
# Table-row order does not determine contingency-table structure.
two[c(6, 1, 4, 2, 5, 3), ] |> freq_test()

reference <- table(mtcars$am, mtcars$cyl)
stopifnot(isTRUE(all.equal(tested$p_fisher[1], fisher.test(reference)$p.value)))
```

When all expected counts exceed 5, only the Pearson p-value is appended:

```{r large-counts}
larger <- mtcars[rep(seq_len(nrow(mtcars)), 10), ]
larger |> group_by(am) |> freq_table(cyl) |> freq_test()
```

This repetition is solely a demonstration of function behavior, not a way
to create independent observations for an actual analysis.

## Generic Headers and Formatting {#pipeline}

Tests depend on counts and category identities, so generic headers and
percentage display give the same p-values. Calculate tests before stacking
results from different outcomes.

```{r pipeline}
mtcars |> group_by(am) |>
  freq_table(cyl, generic_col_names = TRUE, percent = TRUE) |>
  freq_test() |>
  freq_format("percent_group (lcl_group - ucl_group)", name = "estimate", digits = 2) |>
  select(group_01_cat, cat, estimate, p_chi2_pearson, p_fisher)
```

## Missing Values, Empty Cells, and Unsupported Inputs {#limits}

Missing outcomes or group values are categories unless excluded before
`freq_table()`. An unobserved **cell** is filled with zero for the
test even if omitted from the returned table. A completely empty marginal
category makes a two-way test undefined: remove unused levels with
`drop = TRUE`. One-way tests can retain a zero-count category as part
of their equal-probability null.

```{r missing}
d <- data.frame(g = c("a", "a", "b", "b", NA),
                y = c("no", "yes", "yes", NA, "no"))
d |> filter(!is.na(g), !is.na(y)) |>
  group_by(g) |> freq_table(y, drop = TRUE) |> freq_test()
```

There is no automatic test for multiple grouping variables, because a
joint comparison and a test within each stratum answer different questions.
The default method explains this limitation:

```{r unsupported, error=TRUE}
mtcars |> group_by(cyl, vs) |> freq_table(am) |> freq_test()
```
```{r incomplete, error=TRUE}
one |> filter(am == 0) |> freq_test()
```

For statistical details, see R's
[Pearson chi-square](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/chisq.test.html)
and [Fisher exact test](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/fisher.test.html)
documentation. Both rely on an appropriate sampling design; counting
repeated observations does not make them independent.
