---
title: "Using the freq_tbl Function"
author: "Brad Cannell"
date: "Created: 2026-03-19 <br> Updated: `r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    css: styles.css
vignette: >
  %\VignetteIndexEntry{Using the freq_tbl Function}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(dplyr)
devtools::load_all()

data(freq_study)

normalize_tbl <- function(x) {
  x |>
    mutate(across(where(is.numeric), as.numeric))
}

same_tbl <- function(x, y) {
  isTRUE(
    all.equal(
      as.data.frame(normalize_tbl(x)),
      as.data.frame(normalize_tbl(y)),
      check.attributes = FALSE
    )
  )
}
```

`freq_tbl()` is the low-level counting function in `freqtables`. It returns counts and proportions, works naturally inside a `dplyr` pipeline, and respects any grouping created with `group_by()`.

This vignette walks through every main feature of `freq_tbl()` and checks the calculations against explicit `dplyr::count()` workflows. If you want confidence intervals in addition to counts and proportions, use `freq_table()`, which builds on `freq_tbl()`.

# One-way frequency tables

By default, `freq_tbl()` returns the analysis variable, the count (`n`), and the proportion (`prop`).

```{r}
freq_study |>
  freq_tbl(outcome)
```

Missing values are counted by default, which makes the denominator explicit.

```{r}
one_way <- freq_study |>
  freq_tbl(outcome)

one_way_check <- freq_study |>
  count(outcome, .drop = FALSE) |>
  mutate(prop = n / sum(n))

same_tbl(one_way, one_way_check)
```

# Percentages, totals, and display precision

Use `percent = TRUE` to return percentages instead of proportions.

```{r}
freq_study |>
  freq_tbl(outcome, percent = TRUE)
```

```{r}
one_way_percent <- freq_study |>
  freq_tbl(outcome, percent = TRUE)

one_way_percent_check <- freq_study |>
  count(outcome, .drop = FALSE) |>
  mutate(percent = 100 * n / sum(n))

same_tbl(one_way_percent, one_way_percent_check)
```

Use `overall = TRUE` to add the overall denominator.

```{r}
freq_study |>
  freq_tbl(outcome, overall = TRUE)
```

```{r}
one_way_overall <- freq_study |>
  freq_tbl(outcome, overall = TRUE)

one_way_overall_check <- freq_study |>
  count(outcome, .drop = FALSE) |>
  mutate(
    n_total = sum(n),
    prop = n / n_total
  )

same_tbl(one_way_overall, one_way_overall_check)
```

Use `digits` to control how many decimal places are displayed. This is display formatting only; the columns remain numeric and can still be used downstream.

```{r}
digits_tbl <- freq_study |>
  freq_tbl(outcome, digits = 4)

digits_tbl
```

```{r}
is.numeric(digits_tbl$prop)
sum(digits_tbl$prop)
```

# Grouped frequency tables

Grouped analyses are done with `group_by()`, not by passing a second analysis variable to `freq_tbl()`.

```{r}
freq_study |>
  group_by(sex) |>
  freq_tbl(outcome)
```

Within grouped output:

- `n_group` is the denominator within each group.
- `prop_group` is the proportion within each group.

```{r}
grouped_tbl <- freq_study |>
  group_by(sex) |>
  freq_tbl(outcome)

grouped_tbl_check <- freq_study |>
  count(sex, outcome, .drop = FALSE) |>
  group_by(sex) |>
  mutate(
    n_group = sum(n),
    prop_group = n / n_group
  ) |>
  ungroup()

same_tbl(grouped_tbl, grouped_tbl_check)
```

Use `overall = TRUE` to add overall totals to grouped tables, and combine it with `percent = TRUE` if you want percentages on the 0 to 100 scale.

```{r}
freq_study |>
  filter(!is.na(sex), !is.na(outcome)) |>
  group_by(sex) |>
  freq_tbl(outcome, overall = TRUE, percent = TRUE)
```

```{r}
grouped_overall <- freq_study |>
  group_by(sex) |>
  freq_tbl(outcome, overall = TRUE)

grouped_overall_check <- freq_study |>
  count(sex, outcome, .drop = FALSE) |>
  group_by(sex) |>
  mutate(
    n_group = sum(n),
    prop_group = n / n_group
  ) |>
  ungroup() |>
  mutate(
    n_total = sum(n),
    prop_total = n / n_total
  )

same_tbl(grouped_overall, grouped_overall_check)
```

If missing values should not contribute to the denominator, filter them out explicitly before calling `freq_tbl()`.

```{r}
freq_study |>
  filter(!is.na(sex), !is.na(outcome)) |>
  group_by(sex) |>
  freq_tbl(outcome)
```

Trying to pass more than one analysis variable gives an informative error.

```{r, error = TRUE}
freq_study |>
  freq_tbl(sex, outcome)
```

# Generic column names

By default, `freq_tbl()` uses the original data column names. That is usually the most readable output.

When you need to row-bind results across variables, `generic_col_names = TRUE` can be more convenient.

```{r}
freq_study |>
  freq_tbl(outcome, generic_col_names = TRUE)
```

For grouped data, the grouping variables are also returned with generic names.

```{r}
freq_study |>
  group_by(sex) |>
  freq_tbl(outcome, generic_col_names = TRUE)
```

One practical pattern is stacking multiple one-way tables together.

```{r}
bind_rows(
  freq_study |>
    freq_tbl(outcome, generic_col_names = TRUE),
  freq_study |>
    freq_tbl(exposure, generic_col_names = TRUE)
)
```

# Dropping unused factor levels

The `drop` argument is passed to `dplyr::count(.drop = )`.

The `bmi_4cat` variable in `freq_study` includes an unused factor level, `"Extreme Obese"`, which makes it a good example.

With the default `drop = FALSE`, unused levels are retained with `n = 0`.

```{r}
freq_study |>
  freq_tbl(bmi_4cat)
```

With `drop = TRUE`, unused levels are removed.

```{r}
freq_study |>
  freq_tbl(bmi_4cat, drop = TRUE)
```

```{r}
bmi_keep <- freq_study |>
  freq_tbl(bmi_4cat)

bmi_keep_check <- freq_study |>
  count(bmi_4cat, .drop = FALSE) |>
  mutate(prop = n / sum(n))

bmi_drop <- freq_study |>
  freq_tbl(bmi_4cat, drop = TRUE)

bmi_drop_check <- freq_study |>
  count(bmi_4cat, .drop = TRUE) |>
  mutate(prop = n / sum(n))

same_tbl(bmi_keep, bmi_keep_check)
same_tbl(bmi_drop, bmi_drop_check)
```

The same argument works for grouped tables.

```{r}
freq_study |>
  group_by(sex) |>
  freq_tbl(bmi_4cat, drop = TRUE)
```

# Multiple grouping variables

`freq_tbl()` also works with more than one grouping variable. In that case, `n_group` and `prop_group` are calculated within each unique combination of grouping variables.

```{r}
freq_study |>
  filter(!is.na(sex), !is.na(exposure), !is.na(outcome)) |>
  group_by(sex, exposure) |>
  freq_tbl(outcome)
```

```{r}
multi_group_tbl <- freq_study |>
  filter(!is.na(sex), !is.na(exposure), !is.na(outcome)) |>
  group_by(sex, exposure) |>
  freq_tbl(outcome)

multi_group_tbl_check <- freq_study |>
  filter(!is.na(sex), !is.na(exposure), !is.na(outcome)) |>
  count(sex, exposure, outcome, .drop = FALSE) |>
  group_by(sex, exposure) |>
  mutate(
    n_group = sum(n),
    prop_group = n / n_group
  ) |>
  ungroup()

same_tbl(multi_group_tbl, multi_group_tbl_check)
```

# Summary

`freq_tbl()` is most useful when you want a compact table of counts and proportions that fits cleanly into a `dplyr` workflow. The main patterns are:

- Use `group_by()` to define the denominator for grouped proportions.
- Use `filter()` first when missing values should be excluded from the denominator.
- Use `overall = TRUE` when you also want overall totals or proportions.
- Use `percent = TRUE` when you want 0 to 100 scale output.
- Use `generic_col_names = TRUE` when you want to stack results across variables.
- Use `drop = TRUE` when you do not want unused factor levels in the result.

If you need confidence intervals in addition to counts and proportions, move from `freq_tbl()` to `freq_table()`.
